Spontaneous breaking of multipole symmetries

### 2021-12-08 multipoleSSB/ethans_edits.tex
updates from Ethan

\documentclass[pra,aps,twocolumn, amsfonts,amsmath,amssymb,nofootinbib,superscriptaddress]{revtex4}

@@

\newcommand{\note}[1]{\textcolor{red}{#1}}
\newcommand{\outline}[1]{\textcolor{Plum}{#1}}
\newcommand{\todo}[1]{\textcolor{orange}{#1}}
\newcommand{\charlie}[1]{\textcolor{Blue}{#1}}
\newcommand{\RN}[1]{\textcolor{TealBlue}{#1}}
\newcommand{\op}[1]{\mathcal{O}^{(#1)}}
\newcommand{\nn}{\nonumber\\}
\newcommand{\goesto}{\rightarrow}
\renewcommand{\max}{\text{max}}
\newcommand{\eff}{\text{eff}}
\newcommand{\cl}{\text{cl}}
\newcommand{\q}{\text{q}}
\newcommand{\half}{\frac{1}{2}}

@@

\begin{abstract}
Multipole symmetries are of interest both as a window on fracton physics and as a crucial ingredient in realizing new universality classes for quantum dynamics. Here we address the question of whether and when multipole symmetries can be {\it spontaneously broken} in thermal equilibrium. We derive generalized Mermin-Wagner arguments for the total or partial breaking of multipolar symmetry groups and generalized Imry-Ma arguments for the robustness of such multipolar symmetry breaking to disorder. We present both general results and explicit examples. Our results should be directly applicable to quantum dynamics with multipolar symmetries and also provide a useful stepping stone to understanding the robustness of fracton phases to thermal fluctuations, quantum fluctuations, and disorder. 
\end{abstract}

@@

This paper is organized as follows. In Section \ref{multipolegroup} we introduce the multipole group, how to build multipole-invariant field theories, and the generalized Mermin-Wagner argument for total breaking of the maximal multipole symmetry. In Section~\ref{sec:partial} we explore phases in which multipole symmetries are partially broken.
Then, we discuss generalized Imry-Ma arguments in the presence of quenched disorder in Section~\ref{sec:disord}. Finally, we consider an explicit lattice model that illustrates some of these ideas in Section~\ref{sec:example} before concluding with a discussion of open questions in \ref{sec:disc}.

@@

The multipole group is well-explained in Ref.~\cite{Gromov2019}. It generalizes abelian global symmetries such as $\phi(x) \goesto\phi (x) +c$ by allowing a shift by some set of polynomials, $\phi (x) \goesto \phi (x) + \lambda_\alpha P^\alpha(x)$. The variables $\lambda_\alpha$ are symmetry parameters, while $\alpha$ labels the set of polynomials $P^\alpha(x)$. These so-called polynomial shift symmetries~\cite{Griffin2015} all commute with each other. It is helpful to limit ourselves to homogeneous polynomials. We can label these as $P_a^{I_a}$, where $a$ is the degree of the polynomial and $I_a$ is an abstract index that runs over the polynomials of degree $a$.

@@

An example of a multipole group that contains all translations and rotations but is not the maximal multipole group is the group generated by shifts of the form
\begin{align}
\phi \goesto \phi + \lambda_0 P_0^0 + \lambda_{i} P^{i}_1 + \lambda_{I_2} P^{I_2}_2,
\end{align}
where the degree-0 polynomial is $P_0^0=1$. The other polynomials are
\begin{align}
P_1^{i} = x^i,\quad \quad P_2^{I_2} &= \mu^{I_2}_{ij} x^i x^j,
\end{align}
where $\mu^{I_2}$ are a basis for the rank-2 traceless symmetric $d\times d$ matrices. Let us call this group $\mathcal{M}^2_{\text{sym}}$. Recall that the maximal quadrupole group $\mathcal{M}^2_{\text{max}}$ is already only built from symmetric matrices $\mu$. The tracelessness condition thus only removes one polynomial from the set.

@@

We can now consider building field theories that are invariant under various multipole groups. Reference~\cite{Gromov2019} describes the process in detail, so we will only sketch the outline. References~\cite{Griffin2015, Gromov2019} discuss quantum field theories where configurations are weighted by an action. We will also want to consider thermal field theories with Hamiltonian descriptions, so we will abstractly discuss ``kinetic terms" in this section.

The key step is to be able to find spatial derivatives $D$ that annihilate all the polynomials $P_a^{I_a}$ that we want to include in our multipole group. We say that $D$ is order-$s$ if it takes the form
\begin{align}
D = q^i\partial_i + q^{ij}\partial_1\partial_j + \cdots + q^{i_1\dots i_s}\partial_{i_1}\dots\partial_{i_s},
\end{align}
where each $q$ is a symmetric tensor. Although it is not generically possible~\cite{Gromov2019}, we can sometimes find a set of $D_\alpha$ ($\alpha$ is an abstract index) with $s\le a_{\text{max}}$, where $a_\text{max}$ is the highest degree of the polynomials. In this case, the effective field theory will be invariant under some non-maximal multipole symmetry.

Generically, we cannot find any $D$ with $s\le a_\text{max}$ such that $DP_a^{I_a} = 0$ for all the polynomials we want. In that case we can still write down a theory with derivatives 
\begin{align}
D = q^{i_1\dots i_n}\partial_{i_1}\dots \partial_{i_n},
\end{align}
with $n=a_\text{max}+1$. Then, the field theory will be invariant under the maximal multipole group of degree $a_\text{max}$.

Armed with the invariant derivatives $D_\alpha$ ($\alpha$ is another abstract index), we can start to construct a kinetic term. The most general term, at lowest order in derivatives, would be~\cite{Gromov2019}
\begin{align}
K[\phi(x)] &= g_{\alpha\beta} (D_\alpha \phi) (D_\beta \phi).
\end{align}
We will often write the Fourier transform of the kinetic term as $\phi_{-k} K_k \phi_k$.  Requiring some spatial symmetries restricts the choices of $g_{\alpha \beta}$. For the maximal multipole group, enforcing all rotation symmetries results in the kinetic term
\begin{align}
K_a[\phi(x)] &= (\partial_{i_1}\dots \partial_{i_{a+1}} \phi) (\partial_{i_1} \dots \partial_{i_{a+1}} \phi),
\end{align}
which is the kinetic terms studied in Ref.~\cite{Griffin2015}.

@@

\subsection{Generalized Mermin-Wagner argument}

We are now ready to describe the generalized Mermin-Wagner argument for an arbitrary maximal multipole group. This already appears in Ref.~\cite{Griffin2015}, so we are simply reviewing it here in preparation for the more generic cases to follow. Here, we will discuss both classical and quantum systems.

Consider a clean classical system with a spontaneously broken maximal multipole symmetry of degree $a=n-1$.
Then the kinetic term is
\begin{align}
K[\phi(x)] = (\partial_{i_1} \dots \partial_{i_n} \phi)^2, \label{eqn:adisp}
\end{align} 
where $\phi$ is the order parameter field. 
Now assume we are in an ordered phase, but begin to nucleate domains of linear size $L$. Since the order parameter is continuous there will be a thick domain ``wall", in which the order parameter changes gradually from its value in the background phase to its value in the domain. The thickness will be of order $L$.
Equation~\ref{eqn:adisp} tells us the  energy cost of the field changing its value will be minimized if it varies as a polynomial with degree $a+1$. In that case the energy density will scale as $K_k \sim k^{2(a+1)} \sim L^{-2(a+1)}$. Integrating that energy density over a region of size $L$ gives a total energy cost to nucleating domains of $L^{d-2(a+1)}$.

In clean systems there is no energy gain from domain nucleation. When $d\le2(a+1)$, the energy cost is bounded, so the entropy gain favors domain creation. Thus, ordered phases are unstable and spontaneous symmetry breaking cannot occur for $d\le2(a+1)$.

We can reproduce this argument by considering correlation functions. As a warm-up, consider the standard monopole case. Assume there is a monopole U(1) symmetry that is spontaneously broken, leading to a Goldstone boson $\phi(x)$. The correlation function,
\begin{align}
C(x) = \left\langle \phi(x) \phi(0) \right\rangle = \int \frac{d^dk}{(2\pi)^d} \frac{e^{i k \cdot x}}{k^2}, 
\end{align}
diverges when $d\le 2$. Correlation functions should not diverge, so the interpretation is that the Goldstone boson fluctuates strongly enough that the field $\phi(x)$ cannot be well defined. In turn, this tells us the symmetry could not have been broken. This is the classical Mermin-Wagner argument. We should note that, since we only care about long-distance divergences, we don't need to include the complex exponential as long as we only look for divergence at small $k$. We will thus drop this dependence in future calculations.

When we consider a multipole symmetry, the dispersion changes to $K_k=k^{2(a+1)}$. Now the correlation functions will scale as 
\begin{align}
C \sim \int \frac{d^dk}{K_k} \sim \int \frac{d^dk}{k^{2(a+1)}}, \label{eqn:correl}
\end{align}
which diverges for $d\le2(a+1)$. As before, we interpret the divergence of the correlation function as a sign that the symmetry cannot be spontaneously broken. This allows us to reproduce our scaling argument that spontaneous symmetry breaking of the multipole group $\mathcal{M}^a_\max$ cannot occur for $d\le2(a+1)$.

If we look at the quantum (zero temperature) setting, the energy scaling no longer tells us where the critical dimension is. This is because even when there is no energy cost to forming domains, there is no entropy gain. Instead, quantum fluctuations must be the motivation for domain nucleation. To find the quantum critical dimension, we calculate the correlation function,
\begin{align}
C(x) = \left\langle \phi(x) \phi(0) \right\rangle &\sim \int \frac{d\omega d^dk}{\omega^2 + k^{2( a + 1 )}},
\end{align}
which now diverges at $d\le a+1$~\cite{Griffin2015}, where $d$ is the number of spatial dimensions. We have halved the critical dimension, so that spontaneous symmetry breaking cannot occur at $d\le a+1$. It is interesting to note that the dynamical critical exponent in these theories is $z = a+1$, so the classical and quantum critical dimensions are related by $d_\text{cl} = d_\text{q} + z$.

\charlie{Is it more useful to say $d_\text{cl} = d_\text{q} + z$ or $d_\text{cl} = 2 d_\text{q}$, or not try to make any general statement at all?}

@@

The previous analysis suggests a procedure for finding the critical dimension in clean anisotropic systems for breaking the multipole symmetry to the trivial group. Sort each dimension by the degree of its dispersion relation, so that there are $d_n$ dimensions with dispersion $k^{2n}$. Define the effective dimension as 
\begin{align}
d_\eff = \sum_n \frac{d_n}{n}.
\end{align}
In a classical system, spontaneous symmetry breaking cannot occur if $d_\eff\le 2$, while in a quantum system the critical dimension is $d_\eff = 1$.

We should note that the statements in this section were framed in terms of the dispersion of the Goldstone modes rather than the structure of the symmetry group. It is always possible to find the dispersion given the multipole symmetry group, but it may require some basis changes (as in the Haah multipole group).

@@

\subsection{Breaking to a subgroup} \label{sub:subgroup}

In the previous section we assumed we had an order parameter field that fully broke the multipole symmetry. Let us now consider the case where the order parameter breaks the symmetry from the original group $G$ to some subgroup $H$. For simplicity, we will let $G$ be the maximal multipole group $\mathcal{M}^{a}_\text{max}$ and $H$ be $\mathcal{M}^b_\text{max}$, with $b<a$.

The simplest case is when $a=1$ and $b=0$. In physical terms, we are taking a system that conserves total charge and total dipole moment, then condensing objects that carry a dipole moment but no total charge. \note{If $\phi(x)$ is charged under the monopole symmetry, then the operator $\partial_i \phi(x)$ is charged under the dipole part of the symmetry group. \charlie{Do we want correlation functions of $\partial_i\phi$ or $\exp(\partial_i\phi)$?} In a classical system at $T>0$ the correlation functions of this operator are
\begin{align}
\left\langle \partial_i\phi(x) \partial_j\phi(0) \right\rangle &\sim \delta_{ij} \int \frac{d^dk\, k^2}{k^4},
\end{align}
so that the critical dimension is $d=2$, as in the standard Mermin-Wagner theorem. }

@@

It is interesting to note that condensing an object that carries dipole charge must break rotation symmetry. In Sec.~\ref{sec:example} we show how to break the dipole symmetry while preserving a lattice rotation symmetry by considering multiple species of dipole. We are not aware of a way to condense a dipole while preserving continuous rotation symmetry.

\note{In more generality, for this pattern of symmetry breaking there will be a charged operator $\partial_{i_1} \dots \partial{i_m} \phi$, where $m=b+1$. Because $G$ is a maximal multipole group, we can still write the polynomials as in Eqn.~\ref{eqn:basis}. As long as $\phi$ transforms as 
\begin{align}
\delta \phi = \lambda_{I_c} P_c^{I_c}, \label{eqn:partial}
\end{align}
then $\partial_{i_1} \dots \partial_{i_m} \phi$ is invariant under $H=\mathcal{M}^b_\max$. }

>>>>

\documentclass[pra,aps,twocolumn, amsfonts,amsmath,amssymb,nofootinbib,superscriptaddress]{revtex4-2}

@@

\newcommand{\note}[1]{\textcolor{red}{#1}}
\newcommand{\outline}[1]{\textcolor{Plum}{#1}}
\newcommand{\todo}[1]{\textcolor{orange}{#1}}
\newcommand{\charlie}[1]{\textcolor{Blue}{#1}}
\newcommand{\RN}[1]{\textcolor{TealBlue}{#1}}
\newcommand{\op}[1]{\mathcal{O}^{(#1)}}
\newcommand{\nn}{\nonumber\\}
\newcommand{\goesto}{\rightarrow}
\renewcommand{\max}{\text{max}}
\newcommand{\eff}{\text{eff}}
\newcommand{\cl}{\text{cl}}
\newcommand{\q}{\text{q}}
\newcommand{\half}{\frac{1}{2}}
\newcommand{\mmax}[1]{\mathcal{M}^{#1}_\max}

\newcommand{\ethan}[1]{ { \color{blue} \footnotesize \textsf{ethan: \textsl{#1}} }}
\newcommand\be            {\begin{equation}}
\newcommand\ee            {\end{equation}}
\newcommand\ba            {\begin{aligned}}
\newcommand\ea            {\end{aligned}}
\newcommand{\p}{\partial}
\renewcommand{\t}{\theta}
\newcommand{\mcm}{\mathcal{M}}

@@

\begin{abstract}
Multipole symmetries are of interest both as a window on fracton physics and as a crucial ingredient in realizing new universality classes for quantum dynamics. Here we address the question of whether and when multipole symmetries can be spontaneously broken, both in thermal equilibrium and at zero temperature. We derive generalized Mermin-Wagner arguments for the total or partial breaking of multipolar symmetry groups and generalized Imry-Ma arguments for the robustness of such multipolar symmetry breaking to disorder. We present both general results and explicit examples. Our results should be directly applicable to quantum dynamics with multipolar symmetries and also provide a useful stepping stone to understanding the robustness of fracton phases to thermal fluctuations, quantum fluctuations, and disorder. 
\end{abstract}

@@

This paper is organized as follows. In Section \ref{multipolegroup} we introduce the multipole group, and how to build field theories invariant under it. In section \ref{sec:full_breaking} we discuss generalized Mermin-Wagner arguments for situations in which a multipole group is spontaneously broken down to the trivial group, while in Section~\ref{sec:partial} we explore the more subtle case where a nontrivial subgroup is preserved. This analysis allows us to construct examples where a single symmetry-breaking pattern can be described by several distinct types of Goldstone modes, as well as an example where a continuous symmetry is spontaneously broken in one dimension at $T=0$. 
We then discuss generalized Imry-Ma arguments in the presence of quenched disorder in Section~\ref{sec:disord}. Finally, in Section~\ref{sec:example} we consider an explicit lattice model that illustrates some of the ideas of the previous sections. We conclude with a discussion of open questions in \ref{sec:disc}.

@@

The multipole group is well-explained in Ref.~\cite{Gromov2019}. For concreteness, and since we will be interested in situations where a multipolar symmetry group is spontaneously broken, we will describe the multipole group in terms of its action on a compact scalar field $\phi$, which will play the role of a Goldstone boson for the broken multipolar symmetry. We will always imagine $\phi$ as constituting the phase mode for some microscopic order parameter $\psi \sim e^{i\phi}$, with $\psi$ transforming by various $U(1)$ phases under the multipole group. \ethan{are there any examples showing that this interpretation is too restrictive?}

 The multipole group generalizes the internal shift symmetry $\phi(x) \goesto\phi (x) +c$ by allowing a shift by some set of polynomials involving the spatial coordinates, viz. $\phi (x) \goesto \phi (x) + \lambda_\alpha P^\alpha(x)$. The variables $\lambda_\alpha$ are symmetry parameters, while $\alpha$ labels the set of polynomials $P^\alpha(x)$. These so-called polynomial shift symmetries~\cite{Griffin2015} all commute with each other. It is helpful to limit ourselves to homogeneous polynomials. We can label these as $P_a^{I_a}$, where $a$ is the degree of the polynomial and $I_a$ is an abstract index that runs over the polynomials of degree $a$.

@@

An example of a multipole group that contains all translations and rotations but is \emph{not} the maximal multipole group is the group generated by shifts of the form
\begin{align}
\phi \goesto \phi + \lambda_0 P_0^0 + \lambda_{i} P^{i}_1 + \lambda_{I_2} P^{I_2}_2,
\end{align}
where the degree-0 polynomial is $P_0^0=1$. The other polynomials are
\begin{align}
P_1^{i} = x^i,\quad \quad P_2^{I_2} &= \mu^{I_2}_{ij} x^i x^j,
\end{align}
where $\mu^{I_2}$ are a basis for the rank-2 traceless symmetric $d\times d$ matrices. Let us call this group $\mathcal{M}^2_{\text{sym}}$. Recall that the maximal quadrupole group $\mathcal{M}^2_{\text{max}}$ is already only built from symmetric matrices $\mu$. The tracelessness condition thus only removes one polynomial from the set.

@@

We can now consider building IR field theories for phases in which a multipolar symmetry is completely spontaneously broken, with no residual unbroken subgroup. Reference~\cite{Gromov2019} describes the process in detail, so in the following we will be somewhat succinct. 
We will continue to write things down in terms of the compact scalar $\phi$, which (at in the cases for which SSB is not preempted by strong fluctuations) is to be viewed as the Goldstone for the sponatneously broken symmetry. 
% References~\cite{Griffin2015, Gromov2019} discuss quantum field theories where configurations are weighted by an action. We will also want to consider thermal field theories with Hamiltonian descriptions, so we will abstractly discuss ``kinetic terms" in this section.

Since we are only interested in the low-energy physics of the putative symmetry-broken phase, constructing an appropriate IR field theory amounts to nothing more than constructing an appropriate kinetic term for $\phi$. 
To find a kinetic term invariant under a given multipole group $\mathcal{M}$, we need to find operators $D$ built out of spatial derivatives that annihilate all the polynomials in $\mathcal{M}$. If $a_\text{max}$ is the highest degree of the polynomials in $\mathcal{M}$, the simplest derivative operators which generically do the job are of the form \ethan{may or may not want to switch to the composite index notation $q^{I_{a_{\rm max}+1}} \p_{I_{a_{\rm max}+1}}$ employed later}
\begin{align}
D = q^{i_1\dots i_{a_\text{max}+1}}\partial_{i_1}\dots \partial_{i_{a_\text{max}+1}},
\end{align}
for some symmetric tensor $q^{i_1\dots i_{a_\text{max}+1}}$. Although it is not generically possible to do so~\cite{Gromov2019}, we can sometimes find a set of $D_\alpha$ ($\alpha$ is an abstract index) with $s\le a_{\text{max}}$, where $a_\text{max}$ is the highest degree of the polynomials. In this case, the effective field theory will be invariant under some non-maximal multipole symmetry.
% We say that $D$ is order-$s$ if the maximum number of derivatives appearing in $D$ is $s$.

Writing the invariant derivative operators as $D_\alpha$, the most general kinetic term is ~\cite{Gromov2019} 
\begin{align}
K[\phi(x)] &= g_{\alpha\beta} (D_\alpha \phi) (D_\beta \phi)
\end{align}
for some symmetric tensor $g_{\alpha\beta}$.
We will often write the Fourier transform of the kinetic term as $\phi_{-k} K_k \phi_k$.  Requiring some spatial symmetries restricts the choices of $g_{\alpha \beta}$. For the maximal multipole group, enforcing all rotation symmetries results in the kinetic term
\begin{align}
K_a[\phi(x)] &= (\partial_{i_1}\dots \partial_{i_{a+1}} \phi) (\partial_{i_1} \dots \partial_{i_{a+1}} \phi),
\end{align}
which is the kinetic terms studied in Ref.~\cite{Griffin2015}.

@@

\section{Generalized Mermin-Wagner: full multipole breaking}\label{sec:full_breaking}

We are now ready to describe the generalized Mermin-Wagner argument in the case where an arbitrary maximal multipole group is spontaneously broken down to the trivial subgroup. This already appears in Ref.~\cite{Griffin2015}, so we are simply reviewing it here in preparation for the more generic cases to follow. Here, we will discuss both thermal and $T=0$ systems.

Consider first a clean classical system at $T>0$ with a spontaneously broken maximal multipole symmetry of degree $a$.
Then the kinetic term is proportional to 
\begin{align}
K[\phi(x)] = (\partial_{i_1} \dots \partial_{i_{a+1}} \phi)^2. \label{eqn:adisp}
\end{align} 
Now consider nucleating domains of linear size $L$ in which the expectation value of the order parameter --- namely $\langle e^{i\phi}\rangle$ --- varies by an amount of order 1. Since the order parameter is continuous there will be a thick domain ``wall" over which the order parameter changes gradually from its value in the background phase to its value in the domain. The thickness will be of order $L$.
Equation~\ref{eqn:adisp} tells us the  energy cost of the field changing its value will be minimized if it varies as a polynomial with degree $a+1$. In that case the energy density will scale as $K_k \sim k^{2(a+1)} \sim L^{-2(a+1)}$. Integrating that energy density over a region of size $L$ gives a total energy cost to nucleating domains of $L^{d-2(a+1)}$.

In clean systems there is no energy gain from domain nucleation. When $d\le2(a+1)$, the energy cost is bounded, so the entropy gain favors domain creation. Thus, ordered phases are unstable and SSB cannot occur for $d\le2(a+1)$.

We can also reproduce this argument by considering correlation functions. As a warm-up, consider the standard case where a monopole U(1) symmetry is spontaneously broken, leading to a Goldstone boson $\phi(x)$. The correlation function
\begin{align}
C(x) = \left\langle \phi(x) \phi(0) \right\rangle = \int \frac{d^dk}{(2\pi)^d} \frac{e^{i k \cdot x}}{k^2}, 
\end{align}
diverges when $d\le 2$. Correlation functions should not diverge, so the interpretation is that the Goldstone boson fluctuates strongly enough that the field $\phi(x)$ cannot be well defined. In turn, this tells us the symmetry could not have been broken. This is the classical Mermin-Wagner argument. We should note that, since we only care about long-distance divergences, we don't need to include the complex exponential as long as we only look for divergence at small $k$. We will thus drop this dependence in future calculations.

When we consider a (maximal) multipole symmetry, the dispersion changes to $K_k=k^{2(a+1)}$. Now the correlation functions will scale as 
\begin{align}
C \sim \int \frac{d^dk}{K_k} \sim \int \frac{d^dk}{k^{2(a+1)}}, \label{eqn:correl}
\end{align}
which diverges for $d\le2(a+1)$. As before, we interpret the divergence of the correlation function as a sign that the symmetry cannot be spontaneously broken. This allows us to reproduce our scaling argument that SSB of the multipole group $\mathcal{M}^a_\max$ cannot occur for $d\le2(a+1)$.

If we look at the zero temperature setting, the energy scaling no longer tells us where the critical dimension is. This is because even when there is no energy cost to forming domains, there is no entropy gain. Instead, quantum fluctuations must be the motivation for domain nucleation. To find the quantum critical dimension, we calculate the correlation function by way of the (imaginary-time) IR Lagrangian 
\begin{align}
     \mathcal{L}[\phi] = (\partial_\tau\phi)^2 + K[\phi].
\end{align} 
This gives 
\begin{align}
\label{tzerocorr}C(x) = \left\langle \phi(x) \phi(0) \right\rangle &\sim \int \frac{d\omega d^dk}{\omega^2 + k^{2( a + 1 )}},
\end{align}
which now diverges at $d\le a+1$~\cite{Griffin2015}, where $d$ is the number of spatial dimensions. We have halved the critical dimension, so that SSB cannot occur at $d\le a+1$. Note that the dynamical critical exponent in these theories is $z = a+1$, so that as expected, the classical and quantum critical dimensions are related by $d_\text{cl} = d_\text{q} + z$.

@@

The previous analysis suggests a procedure for finding the critical dimension in clean anisotropic systems for breaking the multipole symmetry to the trivial group. First, sort each dimension by the degree of its dispersion relation, so that there are $d_n$ dimensions with dispersion $k^{2n}$. Then define the effective dimension as 
\begin{align}
d_\eff = \sum_n \frac{d_n}{n}.
\end{align}
In a classical system, we then conclude that SSB cannot occur if $d_\eff\le 2$, while in a quantum system the critical dimension is $d_\eff = 1$.

We should emphasize that the statements in this subsection were framed in terms of the dispersion of the Goldstone modes rather than the structure of the symmetry group. It is always possible to find the dispersion given the multipole symmetry group, but it may require some basis changes (as in the Haah multipole group).

@@

In the previous section we studied situations in which a multipole group was fully broken. In this section we turn our attention to examining what happens when the symmetry breaking is incomplete, with a nontrivial subgroup remaining unbroken. A simple example that illustrates why this problem is nontrivial is the following.

Consider a situation in which a dipolar symmetry $\mcm^1_{\rm max}$ is spontaneously broken to its monopolar subgroup $\mcm^0$, and momentarily specialize to the case of $d=3$ and $T>0$. In the spirit of Goldstone's theorem, the most naive thing to do when analyzing the symmetry broken phase would be to write down an action involving a set of $d=3$ Goldstone fields $\theta_j$, with $\t_j$ transforming linearly under the dipolar part of the symmetry group, but remaining invariant under the unbroken monopole subgroup, so that $\theta_j \mapsto \theta_j + \beta_j$ under a transformation parameterized by $P(x) = \alpha + \beta_j x^j$. In this case, the IR field theory for the putative symmetry-breaking phase would be of the form 
\begin{align}
\label{vector} \mathcal{L}= \sum_j  (\partial_\tau \theta_j)^2 + \sum_{i,j} g_{ij} (\partial_i \theta_j)^2.
\end{align}
As we are working in $d=3$ and at $T>0$, the theory \eqref{vector} indeed has SSB for the dipole symmetry, with $e^{i\t_j}$ developing long-range order.

The theory \eqref{vector} is not the only possibility consistent with this pattern of symmetry breaking, however. Indeed, consider instead the scalar field theory 
\begin{align}
\label{singlescalar}    \mathcal{L} = (\partial_\tau\phi)^2 + g_{ij} (\partial_i\partial_j\phi)^2, \end{align}
where $\phi$ transforms under the symmetry as $\phi \mapsto \phi + \alpha + \beta_jx^j$. 
 In the scalar theory \eqref{singlescalar}, the analysis of section \ref{sec:full_breaking} shows that the monopole symmetry is {\it not} spontaneously broken, with $e^{i\phi}$ possessing short-ranged correlation functions. However, the dipolar symmetry does indeed have SSB, and it is easy to check that the operators $e^{i\p_j\phi}$ --- which transform under the symmetry in the same way as $e^{i\t_j}$ --- have long-range order. Thus in this case, the two theories \eqref{vector} and \eqref{singlescalar} exhibit the same pattern of symmetry breaking, despite having a different number of Goldstone modes.\footnote{We will refer to the mode $\phi$ in the theory \eqref{singlescalar} as the ``Goldstone mode'' of the spontaneously broken dipole symmetry, since it is the only mode in the theory. Note however that it is $\p_j\phi$ (and not $\phi$ itself) which transforms under the symmetry action in the way expected of Goldstone bosons.} In order to formulate the correct version of the Mermin-Wagner theorem in the partial breaking case, it will be necessary to understand why this is so.

\subsection{Goldstone counting} \label{sub:subgroup}

That the number of Goldstone modes is not uniquely determined by the pattern of symmetry breaking is of course not in contradiction with Goldstone's theorem, as illustrated e.g. by the example of the ferromagnet vs the antiferromagnet. Indeed, correctly determining the number of Goldstone modes realized in a given situation depends not just on the symmetry breaking pattern, but also on information relating to how the various charge generators act on the ground state, as nicely described in \cite{watanabe2013redundancies}.

A general formal framework for understanding the counting of Goldstone modes in the present context can be developed by following the analysis of \cite{watanabe2013redundancies}, but here we will simply content ourselves with a heuristic physical argument. For simplicity, we will focus on systems in which the total symmetry group $\mcm_{\rm max}^a$ is spontaneously broken down to some nontrivial subgroup.

Let us define the compressibilities $\kappa_c$ by 
\be \label{kappadef} \kappa_c^{I_c} = \frac{d\langle n_c^{I_c}\rangle}{d\mu_c^{I_c}},\ee 
where $n_c$ is the number density of $c$-pole charges, $\mu_c$ is a conjugate chemical potential, and $I_c = \{i_1,\dots,i_c\}$ is a composite index.

Let $b\leq a$ denote the smallest degree of multipole charge with nonzero compressibility, i.e. consider a situation in which $\kappa_d>0$ for all $b\leq d\leq a$, while $\kappa_d=0$ for all $d<b$. Then the minimal IR Lagrangian for the symmetry broken phase looks schematically like 
\be \label{generall} \mathcal{L} = \kappa_b (\partial_\tau \phi_{I_b})^2 + g_{J_{a-b+1}}(\partial_{J_{a-b+1}}\phi_{I_b})^2,\ee 
where 
%$I_b = \{i_1,\dots,i_b\}$ and $J_{a-b+1}=\{j_1,\dots,j_{a-b+1}\}$ are composite indices with 
$\p_{J_{a-b+1}} \equiv \p_{j_1} \cdots \p_{j_{a-b+1}}$, and where $\phi_{I_b}$ is a field that shifts by a polynomial of degree $a-b$ under the action of $\mmax a$. The compressibility appears as the coefficient of the time derivative term since $\p_\tau \phi_{I_b}$ is the momentum conjugate to $\phi_{I_b}$ in the quantum theory, and hence correlation functions of fluctuations in $n_b$ are determined by those of $\p_\tau \phi_{I_b}$.

A necessary condition for $\mmax b$ to be spontaneously broken is that $\kappa_b \neq 0$. Indeed, if $\mmax b$ is spontaneously broken then $n_b$ creates Goldstones when acting on the ground state, thereby ensuring that the density-density correlator appearing in the calculation of \eqref{kappadef} must be nonzero at long wavelengths. \ethan{is this too laconic?}

However, a nonzero $\kappa_b$ does not {\it necessarily} imply that $\mmax b$ is spontaneously broken. $\kappa_b\neq 0$ means that the system possesses gapless modes created by $n_b$, but it is possible for fluctuations to be strong enough so that these modes cannot be associated with the Goldstone modes of a spontaneously broken $\mmax b$. Thus the theory \eqref{generall} does not necessarily describe a phase where $\mmax b$ is spontaneously broken, and the exact pattern of symmetry breaking depends on the spatial dimension and whether or not $T$ is nonzero. For example, we can again consider the theory in \eqref{singlescalar} at $d=3,T>0$: even though the charge compressibility $\kappa_0$ (the coefficient of $(\p_\tau\phi)^2$ in \eqref{singlescalar}) is nonzero, $\phi$ fluctuates strongly enough that the monopole symmetry is not spontaneously broken.

\subsection{Generalized Mermin-Wagner: partial multipole breaking}

In the previous subsection, we saw that the number of Goldstone modes in the symmetry-breaking theory is determined by the smallest degree of multipole charge $b$ such that $\kappa_b\neq0$. We now turn to determining the symmetry breaking pattern that occurs for a given choice of $b,d,$ and $T$, thereby allowing us to formulate a generalized version of the Mermin-Wagner theorem for the case of partial multipole breaking.

Since $\kappa_d=0$ if $d<b$, none of the subgroups $\mmax {d<b}$ can be spontaneously broken. Suppose then that the symmetry is spontaneously broken down to $\mmax {c}$, where $c\geq b$. To find the allowed values of $c$ consistent with this assumption, we need to calculate correlation functions of operators of the form $\exp(i\p_{J_{c-b}}\phi_{I_b})$, these being the operators transforming nontrivially under $\mmax c$ with the longest-range correlation functions. We have 
\be\ba \langle \p_{J_{c-b}}\phi_{I_b}(x) \p_{J_{c-b}}\phi_{I_b}(0)\rangle & \sim \int \frac{d\omega \, dk\, k^{d-1+2(c-b)}}{\omega^2 + k^{2(a-b+1)}} \\ 
& \sim \int dk\, k^{d-1+2(c-b) - \varsigma_T(a-b+1)}, \ea\ee 
where we have defined the function $\varsigma_T$ such that $\varsigma_T = 1$ if $T=0$ and $\varsigma_T = 2$ if $T>0$. Consequently, we find that $\mmax c$ can be spontaneously broken only in dimensions $d$ such that 
\be \label{pbreak_mw} d>2(b-c) + \varsigma_T(a-b+1).\ee 
For a given choice of symmetry breaking pattern (captured by $a$ and $c$) and a given choice of which compressibilities are nonzero (captured by $b$), the quantity on the RHS determines the lower critical dimension for symmetry breaking, thereby generalizing the Mermin-Wagner theorem to the partial breaking case. 
As a sanity check on \eqref{pbreak_mw}, note that when we set $c=b=0$ (full multipole breaking) we recover the lower critical dimension of $\varsigma_T(a+1)$ obtained in the previous section. Also note that as expected, the lower critical dimensions in the quantum $(\varsigma_T = 1)$ and classical $(\varsigma_T = 2)$ cases are related by $d_{\rm cl} = d_{\rm q} + z$, where $z=a-b+1$ is the dynamical exponent of the Goldstone theory \eqref{generall}.

Finally, as an interesting example, consider the case where the dipolar group $(a=1)$ is broken down to the monopole group $(b=0,c=1)$. At $T=0$, we see from \eqref{pbreak_mw} that this is possible provided that $d>0$. Thus the partial breaking of dipole symmetry provides us with an example where a continuous symmetry can be spontaneously broken in $d=1$ at zero temperature.

%In a classical system at $T>0$ the correlation functions of this operator are
%\begin{align}
%\left\langle \partial_i\phi(x) \partial_j\phi(0) \right\rangle &\sim \delta_{ij} \int \frac{d^dk\, k^2}{k^4},
%\end{align}
%so that the critical dimension is $d=2$, as in the standard Mermin-Wagner theorem for monopole symmetries. This allows us to interpret the dipole charges of the monopole symmetry as monopole charges of the dipole part of the symmetry group.
%However, this interpretation does not hold for  quantum systems. 
%The correlation function scales as
%\begin{align}
%\left\langle \partial_i\phi(x) \partial_j\phi(0) \right\rangle &\sim \delta_{ij} \int \frac{d\omega\, d^dk\, k^2}{\omega^2 + k^4}\nn
%&= \delta_{ij} \int \frac{d^dk\, k^2}{k^2},
%\end{align}
%which is always IR-finite. We thus conclude that at $T=0$, $\mcm^1_{\rm max}$ can be spontaneously broken down to $\mcm^0$ in any dimension

@@

%
%
%In more generality, for $G = \mathcal{M}^{a}_\text{max}$ and $H = \mathcal{M}^b_\text{max}$ there will be a charged operator $\partial_{i_1} \dots \partial_{i_{b+1}}\phi$ with a dispersion as in Eqn.~\ref{eqn:adisp}. Because $G$ is a maximal multipole group, we can still write the polynomials as in Eqn.~\ref{eqn:basis}. Recall that $P_c^{I_c}$ is a degree-$c$ monomial. As long as $\phi$ transforms as 
%\begin{align}
%\delta \phi = \lambda_{I_c} P_c^{I_c}, \label{eqn:partial}
%\end{align}
%then $\partial_{i_1} \dots \partial_{i_{b+1}} \phi$ is invariant under $H=\mathcal{M}^b_\max$. 
%
%Classical correlation functions for these operators take the form
%\begin{align}
